Flaring star-planet interactions in Kepler and TESS

### 2023-05-31 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract with filler text
\begin{abstract}
    In many of the star-planet systems discovered so far, the innermost planets orbit within only a few stellar radii. In these systems, the planets become in-situ probes of the extended stellar magnetic field. As they move, they can trigger flares on the stellar surface. Until recently, however, searches for such planet-induced flaring suffered from small sample sizes, and poor orbital phase coverage.

    In this work, we used the excellent phase coverage of the archives of Kepler and the Transiting Exoplanet Survey Satellite to search for star-planet systems with flaring star-planet interactions. We compiled a catalog of all their flares, and searched for planet-induced flaring, i.e., flares clustering in phase with the planet's orbit.

@@

On the one hand, a close-in planet that changes the star's rotation and activity makes it more difficult for us to model the system's evolution, as it affects fundamental stellar parameters. Ages inferred from rotation will be biased, if the star is spun up or spun down relative to stars without close-in companions~\citep{tejadaarevalo2021further, brown2014discrepancies, maxted2015comparison}. Analogously, altered activity levels may mimic an older or younger host than it actually is~\citep{ilic2022tidal}. On the other hand, measuring in what ways and how much close-in planets alter their hosts' behavior can provide important insights into the habitability of the entire system~\citep{shkolnik2018signatures}. Even if they orbit far inside the habitable zone, such planets probe the system's space weather and mass loss by revealing the extent of the Alfv\'en zone through SPI~\citep{kavanagh2021planetinduced, chebly2022destination}. % Prospectively, planet-induced stellar activity could quantify otherwise difficult to measure planetary properties such as the strength and extent of their magnetic field~\citep{cauley2019magnetic}.

@@

\subsection{Flares}
\label{sec:intro:flares}
In this study, we will look at stellar flares as another way of dissipating the energy deposited in the magnetic field. Flares are strong and impulsive eruptions released by reconnection and subsequent relaxation of field lines in the corona~\citep{svestka1976solar,priest2002magnetic}. In contrast to solar flares, stellar flares often are extremely energetic, sometimes enhancing the stellar flux by orders of magnitude~\citep{maehara2012superflares, shibayama2013superflares, paudel2018k2} so that they can be detected in stars over 1kpc away~\citep{chang2015photometric}. But since they occur randomly in time, individual events can be hard to catch. %We can observe flares in nearly all electromagnetic bands -- from radio to X-ray~\citep{benz2010physical}.

@@

If we picture the magnetic field lines connecting the planet to the star in the sub-Alfv\'enic zone, we expect that the stellar footpoint of these lines will move along with the planet. Flares triggered by the planet will occur at a preferred orbital phase, namely whenever the footpoint is visible to the observer. Depending on the system's viewing geometry, i.e., the inclination of the rotation axis, the spin-orbit (mis-)alignment, and the extent and configuration of the large scale magnetic field -- the footpoint could be always in view, or never. In expectation, however, in star-planet systems with favorable viewing geometries and magnetic field strengths and configurations, the footpoint visibility will be modulated in phase with the planetary orbit. The resulting phase preference of planet-induced flares introduces a deviation from a random distribution of flares. Space missions like Kepler and TESS provide light curves that cover all phases of the planet's orbit, so that phase-correlated planet-induced flares can be detected against a background of uncorrelated intrinsic flaring.

@@

\subsection{Custom Anderson-Darling test}
\label{sec:methods:adtest}
For each star-planet system with three or more flares in its light curves, we test for deviations from a random distribution of flares with orbital phase. We use the same customized Anderson-Darling test as in \cite{ilin2022searching}. In brief, we first take the number of flares observed in the TESS and Kepler data to calculate a base flare rate for each light curve. We then calculate how often each orbital phase has been observed, also for each light curve separately. With the phase coverage and base flare rate combined, we can tell how many flares we would expect to see in any given phase bin if flares were randomly distributed. We then aggregate this number of expected flares per time bin into one expected distribution by summing over all available light curves. As a last step, we compare the expected distribution to the observed one using an Anderson-Darling test. We repeat this test with different phase offsets to account for potential biases in sensitivity of the test at different phases.

@@

\begin{eqnarray}
    B &= 199\,\text{G} \cdot R\rm o^{-1.26\pm 0.1} \;(\text{if}\; Ro > 0.13) \\
    B &= 2050\,\text{G} \cdot R\rm o^{-0.11\pm 0.03} \;(\text{if}\; Ro < 0.13) 
\end{eqnarray}

@@

We compared our derived values for $R$o and $B$ with existing estimates of coronal emission from~\citet{foster2022exoplanet}. We show $L_X/L_{\rm bol}$ as a function of $Ro$ and $B$ in Fig.~\ref{fig:lx}. The X-ray emission relation to Rossby number in our sample follows that of stars that are not known to host close-in planets~\citep{wright2011stellaractivityrotation}, spanning both the saturated and unsaturated regime. The X-ray emission and our estimated magnetic field closely follow each other, also consistent with stars not selected for close-in planets~\citep{reiners2022magnetism}. There are two exceptions: GJ 3082 appears underluminous in X-ray compared to its Rossby number and expected magnetic field strength. AU Mic is slightly overluminous, falling somewhat above the saturated level of $L_X/L_{\rm bol}$. However, the star still follows the $B-L_X/L_{\rm bol}$ relation.

@@

We repeated the coherence calculation for $P_{rot}$, and found that the ratio is $>0.1$ for most stars~(Figure~\ref{fig:coherence_hist}, top panel), so that a statistical analysis of rotational flare modulation is not feasible.

\citet{fischer2019timevariable} argue that the synodic period, at which the sub-planetary point crosses the same (magnetic) surface element on the star, may result in a clearer interaction signal, because a magnetically active region that is prone to planet-induced flares, will be passed only once per that period. However, the uncertainty in $P_{\rm rot}$ does not allow us to search for modulation with synodic period. We argue, however, that in a transiting system with high orbital inclination, the effect with synodic period reduces to modulation with the orbital period:

If the orbital axis inclination is high, then the passage will regularly occur on the back of the star relative to the observer. The true period will therefore be additionally modulated by the visibility, resulting in an intermittency of the signal. As a consequence, to detect a modulation with the synodic period, a significantly longer observing baseline is needed than to detect a modulation with the orbital period. Additionally, if there are many active regions (at the same time, or in sequence) that produce flares when the sub-planetary footpoint passes across, the effect with the synodic period further converges to a modulation with orbital period. 
In our sample, the observing baseline is multiple years long for many systems. Most stars, even the most rigidly rotating late M dwarfs, show variability in their spot modulation on such long timescales~\citep[e.g.,][]{giles2017kepler,davenport2015spots, namekata2019lifetimes}).

@@

\subsection{Flare phase distributions}
\label{sec:results:phasedist}
Our custom Anderson-Darling test compares the measured distribution of orbital phases of the flares to the expected distribution, and returns the significance of the deviation between the two. In the expected distribution, the same number of flares would be distributed randomly across all phases, as the overwhelming majority of studies looking for variation with rotational phase suggest~\citep[see, e.g.,][]{doyle2018investigating,howard2021evryflare}. Figures~\ref{fig:cumdist_transiting} and \ref{fig:cumdist_rv} show the flare phase distributions along with the corresponding expected distributions for transiting and non-transiting systems, respectively. The expected distributions usually deviate from a straight line because we take into account the coverage of the orbital phases by the Kepler and TESS observations, as well as the different flare rates in each light curve, which arise due to varying noise levels between individual Quarter/Sectors, and Kepler and TESS.

@@

Table~\ref{tab:maintable_der} lists the $p$-values for each star-planet system. There is no star-planet system for which we find a $>3\sigma$ signal of star-planet interaction in our sample. However, overall, the significance of the deviation increases with higher expected power $P_{xx}$ of interaction, where the subscript denotes different scenarios introduced in Section~\ref{sec:methods:pspi}. Table~\ref{tab:maintable_der} also lists the derived parameters required in these scaling laws, i.e., Rossby number $R$o, surface-average magnetic field strength $B$, and relative velocity between the planet and a co-rotating magnetic field $v_{\rm rel}$.

Figure~\ref{fig:adtest_bp} shows that in every scenario we considered (Alfv\'en wing or stretch-and-break mechanism, with or without planetary magnetic field), there are two branches on the high end of expected powers. On one branch, the significance of the measured interaction is correlated with the expected power $P_{xx}$. On the other branch, no interaction is measured regardless of the expected power $P_{xx}$.

@@

\begin{figure*}[ht!]
    \script{paper_adtest_vs_value_scatterplots.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/PAPER_ADtest_bg.png}
        \caption{
            Expected power of SPI vs. AD test results, assuming four different scenarios, color-coded by stellar surface field strength. Top row: Stretch-and-break scenario. Bottom row: Alfv\'en wing scenario. Left column: assuming $B_p=1\,$G. Right column: assuming $B_p=1\,$G. While the distribution of $p$-values is consistent with no interaction, all scenarios indicate lower $p$-values only for high expected powers of interaction. See Fig.~\ref{fig:adtest_ro} in the Appendix for the same figure, color-coded by Rossby number.
        }
        \label{fig:adtest_bp}
    \end{centering}
\end{figure*}

@@

This relation links the mean AS size with the maximum large-scale radial magnetic field ($B_{\rm r}^{\rm max}$) obtained in Zeeman-Doppler Imaging reconstructions  \textcolor{red}{(references here)}. The relation was derived from 3D~MHD stellar wind simulations cool main sequence stars (F, G, K, and M), which incorporated observed ZDI maps as boundary condition for the stellar radial magnetic field (for more details on the models see Chebly et al. 2023). However, the stellar magnetic field~($B$) in this work (Table XX), derived from the Rossby number of each star (Eq. XX), corresponds to the total, unsigned magnetic field strength on the stellar surface. As such, it is necessary to estimate how $B_{\rm r}^{\rm max}$ relates to $B$.

In \cite{reiners2022magnetism}, it was reported that for Sun-like stars with a total magnetic field strength of up to 2000~G, the longitudinal magnetic field derived from Stokes V spectropolarimetric observations ($B_{\rm V}$) accounts for $\sim$\,$10$\% of the total magnetic field. \textbf{add ref.} showed that this trend also holds for M-dwarfs with magnetic fields up to 4000~G. Therefore, as a first assumption, we consider that the longitudinal magnetic field of our stars would reach 15\% of the total magnetic field (i.e. $B_{\rm V} = 0.15B$), given that most systems in our sample have B~$>$~2000 G.

The second assumption is that $B_{\rm V}$ can be equal to or lower than the maximum radial magnetic field strength. This is because, unlike the maximum radial magnetic field, which is measured at a specific location on the stellar surface, $B_{\rm V}$ may suffer from cancellation as it is measured across a specific area. In this context, the maximum radial magnetic field strength can be considered as the upper limit of $B_{\rm V}$ (15\%~of~B~=~$B_{\rm r}^{\rm max}$). Under these two assumptions, we will be estimating  an upper limit on $\overline{AS}_{R}$.

what we think about it (E and J contributions, comments from others  welcome)
TAP 26 and KOI-12 appear sub-Alfv\'enic in our estimate. So it seems unlikely that the absence of interaction is caused by the planets orbiting outside the Alfv\'en radius.

>>>>

% Abstract with filler text
\begin{abstract}
    In many of the star-planet systems discovered so far, the innermost planets orbit within only a few stellar radii. In these systems, the planets become in-situ probes of the extended stellar magnetic field. As they move, they are expected to trigger flares on the stellar surface. Until recently, however, searches for such planet-induced flaring suffered from small sample sizes, and poor orbital phase coverage.

    In this work, we used the excellent phase coverage of the archives of Kepler and the Transiting Exoplanet Survey Satellite to search for star-planet systems with flaring star-planet interactions. We compiled a catalog of all their flares, and searched for planet-induced flaring, i.e., flares clustering in phase with the planet's orbital phase.

@@

On the one hand, a close-in planet that changes the star's rotation and activity makes it more difficult for us to model the system's evolution, as it affects fundamental stellar parameters. Ages inferred from rotation will be biased, if the star is spun up or spun down relative to stars without close-in companions~\citep{tejadaarevalo2021further, brown2014discrepancies, maxted2015comparison}. Analogously, altered activity levels may mimic an older or younger host than it actually is~\citep{ilic2022tidal}. On the other hand, measuring in what ways and how much close-in planets alter their hosts' behavior can provide important insights into the habitability of the entire system~\citep{shkolnik2018signatures}. Even if they orbit far inside the habitable zone, such planets probe the system's space weather and mass loss~~\citep{cohen2011dynamics, cohen2015interaction, hazra2022impact} by revealing the extent of the Alfv\'en zone through SPI~\citep{kavanagh2021planetinduced, chebly2022destination}. % Prospectively, planet-induced stellar activity could quantify otherwise difficult to measure planetary properties such as the strength and extent of their magnetic field~\citep{cauley2019magnetic}.

@@

\subsection{Flares}
\label{sec:intro:flares}
In this study, we will look at stellar flares as another way of dissipating the energy stored in the magnetic field. Flares are strong and impulsive eruptions released by reconnection and subsequent relaxation of field lines in the corona~\citep{svestka1976solar,priest2002magnetic}. In contrast to solar flares, stellar flares often are extremely energetic, sometimes enhancing the stellar flux by orders of magnitude~\citep{maehara2012superflares, shibayama2013superflares, paudel2018k2} so that they can be detected in stars over 1kpc away~\citep{chang2015photometric}. But since they occur randomly in time, individual events can be hard to catch. %We can observe flares in nearly all electromagnetic bands -- from radio to X-ray~\citep{benz2010physical}.

@@

If we picture the magnetic field lines connecting the planet to the star in the sub-Alfv\'enic zone, we expect that the stellar footpoint of these lines will move along with the planet. Flares triggered by the planet will occur at a preferred orbital phase, namely whenever the footpoint is visible to the observer. Depending on the system's viewing geometry, i.e., the inclination of the rotation axis, the spin-orbit (mis-)alignment, and the extent and configuration of the large scale magnetic field -- the footpoint could be always in view, or never. In expectation, however, in star-planet systems with favorable viewing geometries and magnetic field strengths and configurations, the footpoint visibility will be modulated in phase with the planetary orbit~(see, e.g., Fig.~4 in \citet{strugarek2019chasing} for an illustration XXXX maybe make my own sketch? XXXX). The resulting phase preference of planet-induced flares introduces a deviation from a random distribution of flares. Space missions like Kepler and TESS provide light curves that cover all phases of the planet's orbit, so that phase-correlated planet-induced flares can be detected against a background of uncorrelated intrinsic flaring.

@@

\subsection{Custom Anderson-Darling test}
\label{sec:methods:adtest}
For each star-planet system with three or more flares in its light curves, we test for deviations from a random distribution of flares with orbital phase. We use the same customized Anderson-Darling test as in \cite{ilin2022searching}. In brief, we first take the number of flares observed in the TESS and Kepler data to calculate a base flare rate for each light curve. We then calculate how often each orbital phase has been observed, also for each light curve separately, to ensure that the different detection thresholds for TESS and Kepler do not bias the base rate. With the phase coverage and base flare rate combined, we can tell how many flares we would expect to see in any given phase bin if flares were randomly distributed. We then aggregate this number of expected flares per time bin into one expected distribution by summing over all available light curves. As a last step, we compare the expected distribution to the observed one using an Anderson-Darling test. We repeat this test with different phase offsets to account for potential biases in sensitivity of the test at different phases.

@@

\begin{eqnarray}
    B &= 199\,\text{G} \cdot R\rm o^{-1.26\pm 0.1} \;(\text{if}\; Ro > 0.13) \label{Eq3}\\
    B &= 2050\,\text{G} \cdot R\rm o^{-0.11\pm 0.03} \;(\text{if}\; Ro < 0.13) \label{Eq4}
\end{eqnarray}

@@

We compared our derived values for $R$o and $B$ with existing estimates of coronal emission. We show $L_X/L_{\rm bol}$ obtained from~\citet{foster2022exoplanet} as a function of $Ro$ and $B$ in Fig.~\ref{fig:lx}. The X-ray emission relation to Rossby number in our sample follows that of stars that are not known to host close-in planets~\citep{wright2011stellaractivityrotation}, spanning both the saturated and unsaturated regime. The X-ray emission and our estimated magnetic field closely follow each other, also consistent with stars not selected for close-in planets~\citep{reiners2022magnetism}. There are two exceptions: GJ 3082 appears underluminous in X-ray compared to its Rossby number and expected magnetic field strength. AU Mic is slightly overluminous, falling somewhat above the saturated level of $L_X/L_{\rm bol}$. However, the star still follows the $B-L_X/L_{\rm bol}$ relation.

@@

\citet{fischer2019timevariable} argue that the synodic period, at which the sub-planetary point crosses the same (magnetic) surface element on the star, may result in a clearer interaction signal, because a magnetically active region that is prone to planet-induced flares, will be passed only once per that period. However, we repeated the coherence calculation for $P_{rot}$, and found that the ratio is $>0.1$ for most stars~(Figure~\ref{fig:coherence_hist}, top panel), so that a statistical analysis of rotational flare modulation is not feasible. %We argue, however, that in a transiting system with high orbital inclination, the effect with synodic period reduces to modulation with the orbital period:

%If the orbital axis inclination is high, then the passage will regularly occur on the back of the star relative to the observer. The true period will therefore be additionally modulated by the visibility, resulting in an intermittency of the signal. As a consequence, to detect a modulation with the synodic period, a significantly longer observing baseline is needed than to detect a modulation with the orbital period. Additionally, if there are many active regions (at the same time, or in sequence) that produce flares when the sub-planetary footpoint passes across, the effect with the synodic period further converges to a modulation with orbital period. 
%In our sample, the observing baseline is multiple years long for many systems. Most stars, even the most rigidly rotating late M dwarfs, show variability in their spot modulation on such long timescales~\citep[e.g.,][]{giles2017kepler,davenport2015spots, namekata2019lifetimes}).

@@

\subsection{Flare phase distributions}
\label{sec:results:phasedist}
Our custom Anderson-Darling test compares the measured distribution of orbital phases of the flares to the expected distribution~(see Section~\ref{sec:methods:adtest}), and returns the significance of the deviation between the two. In the expected distribution, the same number of flares would be distributed randomly across all phases, as the overwhelming majority of studies looking for variation with rotational phase suggest~\citep[see, e.g.,][]{doyle2018investigating,howard2021evryflare}. Figures~\ref{fig:cumdist_transiting} and \ref{fig:cumdist_rv} show the flare phase distributions along with the corresponding expected distributions for transiting and non-transiting systems, respectively. The expected distributions usually deviate from a straight line because we take into account the coverage of the orbital phases by the Kepler and TESS observations, as well as the different flare rates in each light curve, which arise due to varying noise levels between individual Quarter/Sectors, and Kepler and TESS.

@@

Table~\ref{tab:maintable_der} lists the $p$-values for each star-planet system. There is no star-planet system for which we find a $>3\sigma$ ($p<0.0027$) signal of star-planet interaction in our sample. However, overall, the significance of the deviation increases with higher expected power $P_{xx}$ of interaction, where the subscript denotes different scenarios introduced in Section~\ref{sec:methods:pspi}. Table~\ref{tab:maintable_der} also lists the derived parameters required in these scaling laws, i.e., Rossby number $R$o, surface-average magnetic field strength $B$, and relative velocity between the planet and a co-rotating magnetic field $v_{\rm rel}$.

Figure~\ref{fig:adtest_bp} shows that in every scenario we considered (Alfv\'en wing or stretch-and-break mechanism, with magnetized or unmagnetized planets), there are two branches on the high end of expected powers. On one branch, the significance of the measured interaction is correlated with the expected power $P_{xx}$. On the other branch, no interaction is measured regardless of the expected power $P_{xx}$.

@@

\begin{figure*}[ht!]
    \script{paper_adtest_vs_value_scatterplots.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/PAPER_ADtest_bg.png}
        \caption{
            Expected power of SPI vs. AD test results, assuming four different scenarios, color-coded by stellar surface field strength. \textbf{Top panels}: stretch-and-break scenario. \textbf{Bottom panels}: Alfv\'en wing scenario. \textbf{Left panels}: assuming a magnetized planet. \textbf{Right panels}: assuming an unmagnetized planet. While the distribution of $p$-values is consistent with no interaction, all scenarios indicate lower $p$-values only for high expected powers of interaction. See Fig.~\ref{fig:adtest_ro} in the Appendix for the same figure, color-coded by Rossby number.
        }
        \label{fig:adtest_bp}
    \end{centering}
\end{figure*}

@@

This relation links the mean AS size with the maximum large-scale radial magnetic field ($B_{\rm r}^{\rm max}$) obtained in Zeeman-Doppler Imaging reconstructions  \textcolor{red}{(references here)}. The relation was derived from 3D~MHD stellar wind simulations cool main sequence stars (F, G, K, and M), which incorporated observed ZDI maps as boundary condition for the stellar radial magnetic field (for more details on the models see Chebly et al. 2023). However, the stellar magnetic field~($B$) in this work (Table~\ref{tab:maintable_der}), derived from the Rossby number of each star (Eq.~\ref{Eq3}, \ref{Eq4}), corresponds to the total, unsigned magnetic field strength on the stellar surface. As such, it is necessary to estimate how $B_{\rm r}^{\rm max}$ could be related to $B$.

In \cite{reiners2022magnetism} it was reported that for Sun-like stars with a total magnetic field strength of up to 2000~G, the longitudinal magnetic field derived from Stokes V spectropolarimetric observations ($B_{\rm V}$) accounts for $\sim$\,$10$\% of the total magnetic field. \textbf{add ref.} showed that this trend also holds for M-dwarfs with magnetic fields up to 4000~G. Therefore, as a first assumption, we consider that the longitudinal magnetic field of our stars would reach 15\% of the total magnetic field (i.e. $B_{\rm V} = 0.15B$), given that most systems in our sample have B~$>$~2000 G.

The second assumption is that $B_{\rm V}$ will match the maximum radial magnetic field in a ZDI reconstruction. Since $B_{\rm V}$ is integrated over the visible surface of the star, this assumption will only hold if for an observing phase containing the strongest large-scale radial magnetic region, the effect from cancellation of opposite polarities is small. As such, in all other rotational phases we expect $B_{\rm V} < B_{\rm r}^{\rm max}$. Therefore, by assuming $B_{\rm V} = B_{\rm r}^{\rm max}$ we are effectively taking an upper limit on the longitudinal magnetic field strength, which will be translated\footnote[3]{Naturally this will also depend on how good our $15$\% match between $B$ and $B_{\rm V}$ is to reality.} in our estimated values of $\overline{AS}_{R}$.

\textcolor{blue}{what we think about it (E and J contributions, comments from others  welcome)
TAP 26 and KOI-12 appear sub-Alfv\'enic in our estimate. So it seems unlikely that the absence of interaction is caused by the planets orbiting outside the Alfv\'en radius.} \textcolor{red}{JDAG: Thats a good point. However, I think being inside the AS is a required but not sufficient condition for the type of SPI that you are looking for. Because you are searching for flares, you need reconnection + particle acceleration (which are the precursors/drivers of the flare emission observed on different wavelengths). While the coronal material is expected to provide the particles (at least in closed-field Active Regions -- if the reconnection takes place in open-field line regions, the plasma density there is usually low which would also yield weak EM signatures as those are typically a function of $\rho$), one of the difficulties I see is that in the star-planet system, the reconnection of the magnetic field is more likely to occur in the planet end rather than on the stellar end (simply because the AR field strength is probably larger than the planetary field). As such, the field topology will be more 'resistant' closer to the star than to the planet, facilitating the re-organization in the latter case compared to the former. Planetary outflows falling back to the star could help to trigger some instabilities, but those would depend on whether or not the planet is losing any significant mass (which we probably don't know for these targets). There's of course more to it than just field strength or accretion, with things like the torsion and helicity of the field among others (which is why the likelihood of a solar AR to host a flare is not only determined by field strength but also by its topology/complexity), but all of that is too complicated or even impossible to estimate here. The bottom line is that there are multiple reasons that could explain why those two stars that lie deep in the sub-Alfv\'enic regime show no signs of flaring SPI which, without a better characterization of each individual system + some 3D modelling, are very hard to ascertain.}
